Race Science in Mainstream Psychology, 1960-2010

### 2025-04-12 paper/paper.qmd
ready?

In @FernandezPintoKnowBetterNot2017, the Tobacco Industry Research Committee/Council for Tobacco Research provides the paradigm for supporting friendly research. @FernandezPintoKnowBetterNot2017 points out that the TIRC/CTR funded "scientific experts in major universities and research institutions" rather than marginal, fringe, or industry-employed scientists. The purpose was "to use 'good' research as a distraction" from the harmful effects of smoking by "fund[ing] research on heredity, infection, nutrition, hormones, nervous tension, and environmental factors" as causes of lung cancer and heart disease (57). This process of "industrial selection" for favorable research [@HolmanExperimentationIndustrialSelection2017] created the misleading impression that these non-tobacco causes were just as important, perhaps even more important, than smoking [compare @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. Similarly, per Laughlin, the Pioneer Fund was explicitly created to promote research on eugenics and race science as just as important, if not more important, than a supposed "equalitarian dogma" and "environmentalist" explanations of racial inequality *[cite]*.

A key aim of the TIRC/CTR was to recruit friendly scientists, "distinguished scientists, whose authority and academic connections would help the industry’s goal of fostering uncertainty about scientific claims" [@FernandezPintoKnowBetterNot2017 57]. One prominent example not discussed by Fernández is statistician Ronald Fisher [@StolleyWhenGeniusErrs1991; @GundleProveThisIndustrys2010], who proposed a series of genetic explanations for the correlation between smoking and lung cancer in the last few years of his life. *[Shockley and Jensen]*

For the current project, we hypothesized that MQ and certain mainstream journals played a key role in another strategy identified by @FernandezPintoKnowBetterNot2017, the "echo chamber effect." We theorize this effect as involving three components: the "margins," the "mainstream," and the "bridge."  @FernandezPintoKnowBetterNot2017 argues that, using sponsored journals and conferences, "the tobacco industry circumvented peer review standards in publication, without compromising its façade of research supporter" (59). These kinds of marginal venues provide a "safe space" in which favorable conceptual frameworks and empirical research (or pseudo-research) could be developed, protected from the destructive critical scrutiny they would receive in the mainstream. Bridge venues are accepted as part of the mainstream, but cite and draw on work from the margins, thereby mainstreaming these marginal ideas, that is, making them appear to be part of the legitimate body of mainstream scientific research.

@@

![Disproportionate presence of Pioneer-funded authors in three focal topics.\
(A) Number of authors (distinguished by family name) for three focal topics from the medium vocabulary, k= 40 model, by year. Mainstream intelligence (topic 22) shows exponential-like increase while MQ race science (topic 7) and race-and-intelligence research (topic 24) remain flat. \
(B) Pioneer-funded authors as a share (%) of all authors publishing in a given topic, by year (thin lines) and 5-year running mean (thick lines). Usually fewer than 15% of authors in a topic are Pioneer-funded. \
(C) Share of papers (%) with at least one Pioneer-funded author, by year (thin lines) and 5-year running mean (thick lines). After about 1985, more than 15% of papers in topic 24 have at least one Pioneer-funded author. \
(D) Comparison of author share (panel B) to paper share (panel C). Each point represents one year. If Pioneer-funded authors had publishing profiles (y-axis) proportionate to their authorship profiles (x-axis), points would be close to the dark line $y = x$. Points above this line indicate that Pioneer-funded authors had a relatively greater publishing presence than would be expected given their authorship profile. Length of thin vertical lines indicates difference between paper share and author share. \
(E) Empirical density plots of mean difference between paper share and author share for 1,000 randomly drawn sets of 15 authors. Bold vertical lines are the mean values for the Pioneer-funded authors (signed length of lines in panel D). Pioneer-funded authors are similar to randomly drawn sets for topics 7 and 22, but very different for topic 24. ](img/15_presence.png){#fig-presence}

*[more details in supplement]*

Post hoc, we were interested in trying to understand the impact that Pioneer might have had as an institution for supporting friendly scientists. (We thank a reviewer for suggesting this further analysis.) Because we did not have citation data, we were unable to examine "impact" in its usual sense of citations. Instead we chose to examine "presence," which we operationalize as the difference between the share of papers (published in a given topic in a given year) with at least one Pioneer-funded author ("paper share") against the share of all authors (again, for papers published in a given topic in a given year; individuated by family name) who were on our list of Pioneer-funded authors ("author share"). Positive values of "presence" would indicate that Pioneer-funded authors publish more (in a given topic) than might be expected.

Because this was a post hoc analysis, we elected to examine "presence" only for the three focal topics for the medium vocabulary, $k=40$ model (fig. @fig-focal). Fig. @fig-presence shows the results of this analysis.

Fig. @fig-presence A shows that topics 7 (MQ-style race science) and 24 (race-and-intelligence) had stable numbers of authors over the scope of the study, while topic 22 (mainstream intelligence) saw an exponential-like increase. Fig. @fig-presence B indicates that Pioneer-funded authors were a relatively high share of authors in topic 24, often comprising more than 10% of all authors in this topic after 1980. In contrast, these authors were almost never more than 10% of all authors in the other two topics.

Fig. @fig-presence C shows that a greater share of papers in topic 24 had Pioneer-funded authors, compared to the other two topics, throughout the scope of the study but especially after about 1980.

Fig. @fig-presence D is a scatterplot of paper share (subfigure C) against author share (subfigure B). Points above the dark diagonal line (length of the thin colored lines) indicate that, in that year-topic combination, Pioneer-funded authors' paper share was greater than their author share, and in this sense had a disproportionate "presence." For topic 7 (MQ-style race science), Pioneer-funded authors frequently had a small, disproportionately *negative* presence. While for topic 22 (mainstream intelligence) Pioneer-funded authors had a small or modest disproportionately positive presence. In contrast, for topic 24 (race-and-intelligence), Pioneer-funded authors had a substantial, disproportionately positive presence, with in some cases differences between paper share and author share of 20 points or more.

To contextualize this measure of "presence," for each topic we used a permutation-style approach, drawing random sets of 15 authors (who ever published in the topic) and calculating paper share and author share in the same way as for Pioneer-funded authors. Fig. @fig-presence E shows the resulting distribution of mean presence (difference between paper share and author share) across 1,000 such permutations, with vertical lines for the observed values for the Pioneer-funded authors. Compared to the permutations, the Pioneer-funded authors do not have a distinctive presence for topics 7 and 22, but are at the rightmost extreme for topic 24.

@@

@@

Using scientometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly*; that a distinct version of race science discourse, centered on race and intelligence, can be identified in these same mainstream journals across the period 1960-2010; and that authors funded by the Pioneer Fund had a disproportionate presence in this latter race science discourse.  These findings indicate that *Mankind Quarterly* (MQ) was a less important venue for late 20th century race science discourse than mainstream psychology journals — especially *Intelligence* (Int) and *Personality and Individual Differences* (PID).  Indeed, until the 1990s, MQ published a very different kind of race science discourse from the race-and-intelligence research published in Int and PID.  During this same period of time, race scientists served in prominent and influential positions in behavior genetics and psychology, including as society presidents and members of journal editorial boards.  In some cases, some race scientists simultaneously maintained active connections to both the mainstream scientific community and white supremacist organizations [@WinstonScientificRacismNorth2020; @JacksonJr.ArthurJensenEvolutionary2022; @SlobodianUnequalMindHow2023].  Thus the emphasis on MQ in race science historiography may have been misplaced. At the same time, the Pioneer Fund may have played a role in promoting race-and-intelligence research in mainstream psychology.

@@

In terms of the margins-mainstream-bridge model of the "echo chamber effect," there was no bridge that carried ideas originating in MQ on the margins to mainstream journals, because the kind of race science discourse that originated in MQ did not spread to the mainstream. At the same time, within the scope of our data, the race-and-intelligence version of race science discourse was always a mainstream phenomenon.

@@

@@

Furthermore, the maintenance of doubt or ignorance about race within psychological science can be achieved not just via the explicit race science research of a small minority of psychologists who seek to create that doubt or ignorance in the first place [cf. @WinstonWhyMainstreamResearch2020], but also through the norms and traditions of the discipline that are engaged in by the average psychological scientist. For example, Roberts and Mortensen analyzed over 1700 articles published between 1995 and 2018 in seven psychology journals, finding that research using White samples was more likely than research using samples of color to be presented in race-neutral language [@roberts2023challenging]. They reported this finding in the context of a larger argument about how psychology as a discipline treats White as a default and nonracial description of humanity and offering suggestions for how psychology can move past this longstanding implicit yet discipline-wide framework. The treatment of Whiteness as racially neutral by psychology can, like the reluctance of psychological science organizations to reject racial categories as biologically meaningful, further the goals of those who wish to create racialized ignorance, and it is the responsibility of psychological scientists to rebuke efforts that create ignorance.

@@

>>>>

In @FernandezPintoKnowBetterNot2017, the paradigm for supporting friendly research is the Tobacco Industry Research Committee, later known as the Council for Tobacco Research. @FernandezPintoKnowBetterNot2017 points out that the TIRC/CTR funded "scientific experts in major universities and research institutions" rather than marginal, fringe, or industry-employed scientists. The purpose was "to use 'good' research as a distraction" from the harmful effects of smoking by "fund[ing] research on heredity, infection, nutrition, hormones, nervous tension, and environmental factors" as causes of lung cancer and heart disease (57). This process of "industrial selection" for favorable research [@HolmanExperimentationIndustrialSelection2017] created the misleading impression that these non-tobacco causes were just as important, perhaps even more important, than smoking [compare @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. Similarly, per Laughlin, the Pioneer Fund was explicitly created to promote research on eugenics and race science as just as important, if not more important, than non-biological ("environmentalist") explanations of racial social inequality.

For the current project, we hypothesized that MQ and certain mainstream journals played a key role in another strategy identified by @FernandezPintoKnowBetterNot2017, the "echo chamber effect." We theorize this effect as involving three components: the "margins," the "mainstream," and the "bridge" connecting them.  @FernandezPintoKnowBetterNot2017 argues that, using sponsored journals and conferences, "the tobacco industry circumvented peer review standards in publication, without compromising its façade of research supporter" (59). These kinds of marginal venues provide a "safe space" in which favorable conceptual frameworks and empirical research (or pseudo-research) could be developed, protected from the destructive critical scrutiny they would receive in the mainstream. Bridge venues are accepted as part of the mainstream, but cite and draw on work from the margins, thereby mainstreaming these marginal ideas, that is, making them appear to be part of the legitimate body of mainstream scientific research. Where supporting friendly research is a strategy more-or-less entirely within the mainstream of the scientific community, the "echo chamber" strategy is located on the margins or fringes.

@@

@@

![Disproportionate presence of Pioneer-funded authors in three focal topics.\
(A) Number of authors for three focal topics from the medium vocabulary, k= 40 model, by year. Mainstream intelligence (topic 22) shows exponential-like increase while MQ race science (topic 7) and race-and-intelligence research (topic 24) remain flat. \
(B) Pioneer-funded authors as a share (%) of all authors publishing in a given topic, by year (thin lines) and 5-year running mean (thick lines). Usually fewer than 15% of authors in a topic are Pioneer-funded. \
(C) Share of papers (%) with at least one Pioneer-funded author, by year (thin lines) and 5-year running mean (thick lines). After about 1985, more than 15% of papers in topic 24 have at least one Pioneer-funded author. \
(D) Comparison of author share (panel B) to paper share (panel C). Each point is one year-topic combination. Points above the dark line $y = x$ indicate Pioneer-funded authors had a greater paper share than would be expected based on author share. Length of thin vertical lines measures "presence." \
(E) Empirical density plots of "presence," mean difference between paper share and author share for 1,000 randomly drawn sets of 15 authors. Bold vertical lines are the mean values for the Pioneer-funded authors (signed length of lines in panel D). Pioneer-funded authors are similar to randomly drawn sets for topics 7 and 22, but very different for topic 24. ](img/15_presence.png){#fig-presence}

Post hoc, we were interested in trying to understand the impact that Pioneer might have had as an institution for supporting friendly scientists. (We thank a reviewer for comments that suggested this further analysis.) Because we did not have citation data, we were unable to examine "impact" in its usual sense of citations. Instead we chose to examine "presence," which we operationalized as the difference between the share of papers with at least one Pioneer-funded author ("paper share") against the share of all authors who were on our list of Pioneer-funded authors ("author share"). Positive values of "presence" would indicate that Pioneer-funded authors publish more (in a given topic) than might be expected.

Because this was a post hoc analysis, we elected to examine "presence" only for the three focal topics for the medium vocabulary, $k=40$ model (figure @fig-focal). Figure @fig-presence shows the results of this analysis. Details of the methods used here are reported in @sec-presence.

Subfigure @fig-presence A shows that topics 7 (MQ-style race science) and 24 (race-and-intelligence) had stable numbers of authors over the scope of the study, while topic 22 (mainstream intelligence) saw an exponential-like increase. Fig. @fig-presence B indicates that Pioneer-funded authors were a relatively high share of authors in topic 24, often comprising more than 10% of all authors in this topic after 1980. In contrast, these authors were almost never more than 10% of all authors in the other two topics.

Subfigure @fig-presence C shows that a greater share of papers in topic 24 had Pioneer-funded authors, compared to the other two topics, throughout the scope of the study but especially after about 1980.

Subfigure @fig-presence D is a scatterplot of paper share (subfigure C) against author share (subfigure B), by year and topic. The dark diagonal line $y = x$ corresponds to equality between paper share and author share. Distance above (below) this line indicates that Pioneer-funded authors had disproportionately greater (less) "presence" than would be expected based on their authorship profile. For topics 7 (MQ-style race science) and 22 (mainstream intelligence), Pioneer-funded authors consistently had a small negative to modestly positive presence. In contrast, for topic 24 (race-and-intelligence), Pioneer-funded authors consistently had a substantial, disproportionately positive presence, with in some cases differences between paper share and author share of 20 points or more.

Subfigure @fig-presence E uses a resampling method to compare the observed value of presence for the Pioneer-funded authors (dark vertical lines) to a "random" set of other authors publishing in the topic (density curves and rugs). Compared to other authors, the Pioneer-funded authors do not have a distinctive presence for topics 7 and 22, but are at the rightmost extreme for topic 24.

@@

Using scientometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, much more than in *Mankind Quarterly* (MQ); that a distinct version of race science discourse, centered on race and intelligence, can be identified in these same mainstream journals across the period 1960-2010; and that authors funded by the Pioneer Fund had a disproportionate presence in this latter race science discourse, and not in the variety of race science discourse published by MQ.  These findings indicate that MQ was a less important venue for late 20th century race science discourse than mainstream psychology journals — especially *Intelligence* (Int) and *Personality and Individual Differences* (PID).  Indeed, until the 1990s, MQ published a very different kind of race science discourse from the race-and-intelligence research published in Int and PID.  These results contrast with much of the qualitative historiography of scientific racism, which has often emphasized MQ [@MehlerFoundationFascismNew1989; @WinstonScienceServiceFar1998; @TuckerFundingScientificRacism2002; @SchafferScientificRacismAgain2007; @WinstonScientificRacismNorth2020; @SainiDraperMillionsPhilanthropic2022; @AdamsMisappropriationBiologicalAnthropology2021 also emphasize MQ, but in relation to anthropology rather than psychology].

During this same period of time, race scientists served in prominent and influential positions in behavior genetics and psychology, including as society presidents and members of journal editorial boards.  In some cases, some race scientists simultaneously maintained active connections to both the mainstream scientific community and white supremacist organizations [@WinstonScientificRacismNorth2020; @JacksonJr.ArthurJensenEvolutionary2022; @SlobodianUnequalMindHow2023].  Thus the emphasis on MQ in race science historiography may have been misplaced. At the same time, our findings indicate that the Pioneer Fund may have played an important role in promoting race-and-intelligence research in mainstream psychology.

@@

In terms of the margins-mainstream-bridge model of the "echo chamber effect," there was no bridge that carried ideas originating in MQ on the margins to mainstream journals, because the kind of race science discourse that originated in MQ did not spread to the mainstream. At the same time, within the scope of our data, the race-and-intelligence version of race science discourse was always a mainstream phenomenon, and as a mainstream phenomenon was potentially supported by the Pioneer Fund. All together, the Pioneer Fund, *Intelligence*, and *Personality and Individual Differences* appear to fit with the strategy of promoting friendly research, rather than the echo chamber strategy [@FernandezPintoKnowBetterNot2017].

@@

Furthermore, doubt or ignorance about race within the community of psychological science is maintained not only by the Pioneer Fund and its small group of "friendly scientists" [@WinstonWhyMainstreamResearch2020] but also by the ordinary norms and standards of the broader scientific community. @roberts2023challenging analyzed over 1700 articles published between 1995 and 2018 in seven psychology journals, finding that research using White samples was more likely to be presented in race-neutral language than research using samples of people of color. @roberts2023challenging further argued that psychology as a discipline treats White as a default and "nonracial" description of humanity.

@@

<!-- Furthermore, the maintenance of doubt or ignorance about race within psychological science can be achieved not just via the explicit race science research of a small minority of psychologists who seek to create that doubt or ignorance in the first place [cf. @WinstonWhyMainstreamResearch2020], but also through the norms and traditions of the discipline that are engaged in by the average psychological scientist. For example, Roberts and Mortensen analyzed over 1700 articles published between 1995 and 2018 in seven psychology journals, finding that research using White samples was more likely than research using samples of color to be presented in race-neutral language [@roberts2023challenging]. They reported this finding in the context of a larger argument about how psychology as a discipline treats White as a default and nonracial description of humanity and offering suggestions for how psychology can move past this longstanding implicit yet discipline-wide framework. The treatment of Whiteness as racially neutral by psychology can, like the reluctance of psychological science organizations to reject racial categories as biologically meaningful, further the goals of those who wish to create racialized ignorance, and it is the responsibility of psychological scientists to rebuke efforts that create ignorance. -->

@@

# Supplemental analysis of "presence" of Pioneer-funded authors {#sec-presence}

For this supplemental analysis, we focused on the medium vocabulary, $k = 40$ and the same focal topics as in figure @fig-focal. We characterized a paper as "in" one of these topics based on its maximum value of $\gamma$, the topic-document distribution.

## Author name handling and ambiguity error

The Crossref API does not standardize or disambiguate author names. We manually disambiguated the names of the 15 Pioneer-funded authors, e.g., searching the author field for every occurrence of "Rushton" and then identifying all the variations for J. Philippe Rushton. A simple crosswalk table allowed us to standardize all of these variations to a "canonical" version of the name of each Pioneer-funded author, e.g., J.P. Rushton would be replaced with the "canonical" version J. Philippe Rushton before proceeding with further analysis. Fig. @fig-pf-count shows the number of Pioneer-funded authors who published in the three focal topics, by year. Across the entire scope of the data, 12 of the 15 authors ever published in any of these three topics, with Rushton, Lynn, Jensen, Humphreys, and Vernon each publishing 10 or more papers (aggregated across topics).

![Count of Pioneer-funded authors publishing in three focal topics, by year](img/15_pf_count.png){#fig-pf-count}

Names of other authors were not disambiguated, and so, e.g., D.J. Hicks and Daniel J. Hicks would be treated as two different authors for the purposes of this analysis. As a consequence, the total number of authors is probably too high in the estimate of `author_share`, and so the estimate of `author_share` is probably too low. Assuming no issues with the manual disambiguation of Pioneer-funded authors, `paper_share` does not have this problem. Therefore the estimated difference `paper_share - author_share` is probably larger than the true difference. In terms of subfigure #fig-presence E, the vertical bar is probably too far to the right.

This error is unlikely to have significant implications for topics 7 (MQ-style race science) and 22 (mainstream intelligence research). For topic 24 (race-and-intelligence), we judge that a significant error would be 5 percentage points or more, i.e., rather than the estimated mean difference of about 11%, the true mean difference would be about 6% or less. Algebraically,

$$ \frac{p}{t} - \frac{p}{r} = \frac{p}{r - \varepsilon} - \frac{p}{r} > \alpha $$

where $p$ is the count of Pioneer-funded authors; $t$ is the true author count, i.e., after perfect disambiguation; $r$ is the "raw" author count estimated using no disambiguation (except for the Pioneer-funded authors); $\varepsilon$ is the "ambiguity error" $r - t$; and $\alpha$ is the threshold .05 for a significant error in the conclusion.  Solving for $\varepsilon$ gives

$$\varepsilon > r - \frac{p}{\alpha + p/r}$$

To estimate what size of ambiguity error $\varepsilon$ would be required, we first note from @fig-pf-count that at most 5 Pioneer-funded authors are active in topic 24 in any given year. And subfigure @fig-presence A indicates about 20-25 authors are active in this topic each year from the mid 1970s on. Setting $p = 5$ and $r = 20$ gives us $\varepsilon > 3.33$; $p = 5$ and $r = 25$ gives $\varepsilon > 5$; and $p = 3, r = 20$ also gives $\varepsilon > 5$. That is, a disambiguation error of 4-5 authors could be enough to undermine our finding for topic 24.

We therefore manually inspected the list of authors for topic 24 in 1989, 1990, and 1991, to gauge the potential magnitude of ambiguity error. In 1989, topic 24 had 21 distinct authors, including one paper without an author name, and no author name variations that would require disambiguation. In 1990, topic 24 had 24 authors, 1 missing, and no variations. And in 1991 there were 30 authors, none missing, no variations.

For the purposes of this ad hoc analysis, we are satisfied that ambiguity error is probably sufficiently small.

## Paper share and presence

Paper share was calculated simply by asking whether any (at least one) of a given paper's authors appeared on our list of Pioneer-funded authors. This resulted in a dichotomous (yes/no) variable at the paper level, and paper share was calculated as the fraction of papers with a "yes" value.

In subfigure @fig-presence D, years when Pioneer-funded authors had 0 author share, i.e., none of these authors published in the topic, were excluded from calculating the mean value of presence.

## Resampling to contextualize presence

To contextualize this measure of "presence" within topics, we first identified all distinct authors (Pioneer-funded or otherwise, again without disambiguation for the others) across all papers in the given topic. We then used a permutation-style resampling approach: drawing a set of 15 authors and calculating paper share, author share, their difference, and the mean of this difference, exactly as with the Pioneer-funded authors. (In particular, years where author share was 0, i.e., the authors had no publications in the topic, were excluded from calculating the mean difference.)

This approach is equivalent to randomly assigning group labels in a permutation test, and can be interpreted as comparing the presence of Pioneer-funded authors to a "random" set of authors.  We conducted 1,000 resamples for each of the three focal topics.
